---
title: "Extrapolation to tracts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extrapolation to tracts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r setup}
library(CDCAtlas)
library(dplyr)
library(ggplot2)
```

## Overview

CDC AtlasPlus reports many outcomes at the county level. For some analyses,
especially geographic access analyses, we may want estimates at a smaller
geographic unit such as the Census tract.

`CDCAtlas` includes helper functions for extrapolating county-level AtlasPlus
counts to Census tracts using tract-level Census denominators.

The basic idea is:

1. Retrieve county-level CDC AtlasPlus counts.
2. Retrieve tract-level Census population denominators.
3. Match each tract to its parent county.
4. Allocate each county's reported cases across tracts in proportion to the
   tract's share of the relevant county population.

For example, if a tract contains 2% of a county's population in the relevant
denominator, it receives 2% of the county's reported cases.

This is a **population-weighted extrapolation**, not a direct CDC tract-level
surveillance estimate.

## Census data requirements

Tract-level extrapolation requires tract-level population denominators from the
U.S. Census Bureau. `CDCAtlas` uses the `tidycensus` package to retrieve these
denominators.

If you only use `CDCAtlas` to retrieve AtlasPlus county, state, or national data,
you do not need to set up `tidycensus`. However, if you use
`extrapolate_to_tract = TRUE`, you will need:

1. the `tidycensus` package installed
2. a Census API key
3. the API key saved in your R environment

You can install `tidycensus` from CRAN:

```{r, eval=FALSE}
install.packages("tidycensus")
```

Then request a free Census API key from the U.S. Census Bureau and install it
using `tidycensus::census_api_key()`.

See [https://walker-data.com/tidycensus/](https://walker-data.com/tidycensus/) for further details.

## Important assumptions

This method assumes that, within each county and stratum, cases are distributed
across tracts in proportion to the relevant Census denominator.

For example:

- unstratified HIV prevalence may be allocated using total tract population
- sex-stratified counts may be allocated using tract population by sex
- age-stratified counts may be allocated using tract population by age
- race/ethnicity-stratified counts may be allocated using tract population by
  race/ethnicity

This can be useful for geographic access models, but it should not be interpreted
as observed tract-level surveillance data.

## Basic unstratified extrapolation

The simplest use case is extrapolating county-level data to tracts using total
tract population.

```{r}
hiv_tract <- get_atlas(
  disease = "hiv",
  year = 2022,
  geography = "county",
  extrapolate_to_tract = TRUE
)
```

The returned data contain one row per tract, with the county-level AtlasPlus
count allocated to tracts.

```{r}
head(hiv_tract)
```

A typical output includes columns such as:

```{r}
names(hiv_tract)
```

You should expect columns identifying:

- the tract
- the parent county
- the AtlasPlus disease and year
- the county-level cases
- the tract denominator
- the county denominator
- the tract share of the county denominator
- the extrapolated tract cases

The key calculation is:

```r
tract_cases = county_cases * tract_population / county_population
```

or, more generally:

```r
tract_cases = county_cases * tract_denominator / county_denominator
```

## Checking county totals

After extrapolation, tract-level estimates should sum back to the original
county-level counts, allowing for small floating-point differences.

```{r}
hiv_tract %>%
  group_by(county_fips, indicator) %>%
  summarize(
    county_cases = first(county_cases),
    tract_cases_sum = sum(tract_cases, na.rm = TRUE),
    difference = tract_cases_sum - county_cases,
    .groups = "drop"
  ) %>%
  arrange(desc(abs(difference))) %>%
  head()
```

In a clean allocation, the `difference` column should be very close to zero.

## Sex-stratified extrapolation

For sex-stratified AtlasPlus data, the allocation denominator should also be
sex-specific.

```{r}
hiv_tract_sex <- get_atlas(
  disease = "hiv",
  year = 2022,
  geography = "county",
  stratify_by = "sex",
  extrapolate_to_tract = TRUE
)
```

This means that male county cases are allocated across tracts using the male
population in each tract, and female county cases are allocated using the female
population in each tract.

Conceptually:

$$
\widehat{\text{tract cases}}_{\text{male}} =
\text{county cases}_{\text{male}} \times
\frac{\text{tract population}_{\text{male}}}
{\text{county population}_{\text{male}}}
$$

$$
\widehat{\text{tract cases}}_{\text{female}} =
\text{county cases}_{\text{female}} \times
\frac{\text{tract population}_{\text{female}}}
{\text{county population}_{\text{female}}}
$$
The resulting data should include one row per tract per sex stratum.

```{r}
hiv_tract_sex %>%
  count(sex, indicator)
```

You can verify the allocation within each county and sex stratum.

```{r}
hiv_tract_sex %>%
  group_by(county_fips, sex, indicator) %>%
  summarize(
    county_cases = first(county_cases),
    tract_cases_sum = sum(tract_cases, na.rm = TRUE),
    difference = tract_cases_sum - county_cases,
    .groups = "drop"
  ) %>%
  arrange(desc(abs(difference))) %>%
  head()
```

## Age-stratified extrapolation

Age-stratified extrapolation works the same way, except that the Census
denominator is based on tract population within the corresponding AtlasPlus age
group.

```{r}
hiv_tract_age <- get_atlas(
  disease = "hiv",
  year = 2022,
  geography = "county",
  stratify_by = "age",
  extrapolate_to_tract = TRUE
)
```

The output contains one row per tract per age stratum.

```{r}
hiv_tract_age %>%
  count(age, indicator)
```

Again, the diagnostic check is whether tract estimates sum back to the original
county count within each county-age stratum.

```{r}
hiv_tract_age %>%
  group_by(county_fips, age, indicator) %>%
  summarize(
    county_cases = first(county_cases),
    tract_cases_sum = sum(tract_cases, na.rm = TRUE),
    difference = tract_cases_sum - county_cases,
    .groups = "drop"
  ) %>%
  arrange(desc(abs(difference))) %>%
  head()
```

## Race/ethnicity-stratified extrapolation

CDC AtlasPlus commonly uses a combined race/ethnicity variable. For example,
Hispanic/Latino is represented as a single ethnicity category, while the
remaining groups are usually non-Hispanic race categories.

```{r}
hiv_tract_race <- get_atlas(
  disease = "hiv",
  year = 2022,
  geography = "county",
  stratify_by = "race",
  extrapolate_to_tract = TRUE
)
```

Race/ethnicity-stratified extrapolation requires care because Census and
AtlasPlus race/ethnicity categories may not always align perfectly.

The package attempts to use Census variables that correspond to the AtlasPlus
combined race/ethnicity categories, but users should inspect the denominator
mapping before interpreting results.

```{r}
hiv_tract_race %>%
  count(race_ethnicity, indicator)
```

## Multiple stratification variables

Currently not supported

## Working with geometry

Currently not supported

