---
title: "Getting started with CDCAtlas"
output:
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with CDCAtlas}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


**CDCAtlas** provides R functions for retrieving public health surveillance data
from CDC AtlasPlus. This vignette introduces installation, basic queries, and data handling.

## Installation


```{r, eval=FALSE}
# install.packages("remotes")
remotes::install_github("VagishHemmige/CDCAtlas")
```


## Example: State-level chlamydia data

```{r}

library(CDCAtlas)
df <- get_atlas(
  disease   = "chlamydia",
  geography = "state",
  year      = 2022
)

head(df)
```

## Simple visualization

```{r}
library(ggplot2)

ggplot(df, aes(x = geography, y = rate100000)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Chlamydia Rate by State (2022)",
    x = "State",
    y = "Rate per 100,000"
  )
```

## Next steps

- Try different diseases or geographies
- See `?get_atlas` for full argument documentation
- More vignettes coming soon!
