---
title: "Getting started with CDCAtlas"
output:
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with CDCAtlas}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

**CDCAtlas** provides R functions for retrieving public health surveillance data
from CDC AtlasPlus.

This vignette introduces common use patterns:

- retrieving state- and county-level data
- querying different diseases
- using sex, age, and race/ethnicity stratification
- comparing rates across groups
- preparing AtlasPlus data for plotting and analysis

This vignette does **not** cover tract-level extrapolation. That is an advanced
use case covered separately.

## Installation

```{r, eval=FALSE}
# install.packages("remotes")
remotes::install_github("VagishHemmige/CDCAtlas")
```

## Setup

```{r}
library(CDCAtlas)
library(dplyr)
library(ggplot2)
```

## Basic query: state-level chlamydia data

A simple query retrieves one disease, one geography level, and one year.

```{r}
chlamydia_state <- get_atlas(
  disease   = "chlamydia",
  geography = "state",
  year      = 2022
)

head(chlamydia_state)
```

The returned object is a data frame that can be used directly with tidyverse
tools.

```{r}
glimpse(chlamydia_state)
```

## Plot state-level rates

```{r}
chlamydia_state %>%
  filter(!is.na(rate100000)) %>%
  ggplot(aes(x = reorder(geography, rate100000), y = rate100000)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Chlamydia rate by state, 2022",
    x = NULL,
    y = "Rate per 100,000"
  )
```

## Query county-level data

AtlasPlus can also return county-level data for supported diseases.

```{r}
chlamydia_county <- get_atlas(
  disease   = "chlamydia",
  geography = "county",
  year      = 2022
)

head(chlamydia_county)
```

A common workflow is to identify counties with the highest reported rates.

```{r}
chlamydia_county %>%
  filter(!is.na(rate100000)) %>%
  arrange(desc(rate100000)) %>%
  select(geography, cases, rate100000) %>%
  head(10)
```

## Stratification by sex

Many AtlasPlus endpoints support stratification. For example, we can retrieve
state-level gonorrhea data by sex.

```{r}
gonorrhea_sex <- get_atlas(
  disease   = "gonorrhea",
  geography = "state",
  year      = 2022,
  stratify_by = "sex"
)

head(gonorrhea_sex)
```

Now we can compare rates by sex within each state.

```{r}
gonorrhea_sex %>%
  filter(!is.na(rate100000)) %>%
  group_by(sex) %>%
  slice_max(rate100000, n = 10, with_ties = FALSE) %>%
  ungroup() %>%
  ggplot(aes(x = reorder(geography, rate100000), y = rate100000)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~ sex, scales = "free_y") +
  labs(
    title = "States with highest gonorrhea rates by sex, 2022",
    x = NULL,
    y = "Rate per 100,000"
  )
```

## Stratification by age

AtlasPlus also supports age-stratified queries for many disease/geography
combinations.

```{r}
hiv_age <- get_atlas(
  disease   = "hiv",
  geography = "state",
  year      = 2022,
  stratify_by = "age"

)

head(hiv_age)
```

We can summarize national patterns by aggregating across states.

```{r}
hiv_age_summary <- hiv_age %>%
  group_by(age, indicator) %>%
  summarize(
    cases = sum(cases, na.rm = TRUE),
    .groups = "drop"
  )

hiv_age_summary
```

```{r}
hiv_age_summary%>%
  filter(indicator=="HIV diagnoses")%>%
  ggplot(mapping=aes(x = age, y = cases)) +
  geom_col() +
  labs(
    title = "HIV diagnoses by age group, 2022",
    x = "Age group",
    y = "Cases"
  )
```

## Stratification by race/ethnicity

CDC AtlasPlus commonly uses a combined race/ethnicity variable rather than
separate race and Hispanic ethnicity fields.

```{r}
hiv_race <- get_atlas(
  disease        = "hiv",
  geography      = "state",
  year           = 2022,
  stratify_by = "race"

)

head(hiv_race)
```

A useful first look is to compare total reported cases by race/ethnicity.

```{r}
hiv_race_summary <- hiv_race %>%
  group_by(race_ethnicity, indicator) %>%
  summarize(
    cases = sum(cases, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(cases))

hiv_race_summary
```

```{r}

hiv_race_summary%>%
  filter(indicator=="HIV diagnoses")%>%
  ggplot(mapping =aes(x = reorder(race_ethnicity, cases), y = cases)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "HIV cases by race/ethnicity, 2022",
    x = NULL,
    y = "Cases"
  )
```

## Multiple stratification variables

You can request combinations of stratification variables when supported by the
underlying AtlasPlus endpoint. For example, HIV data can be queried by both sex
and age.

```{r}
hiv_age_sex <- get_atlas(
  disease   = "hiv",
  geography = "state",
  year      = 2022,
  stratify_by = c("sex", "age")

)

head(hiv_age_sex)
```

This allows direct comparison of age distributions by sex.

```{r}
hiv_age_sex_summary <- hiv_age_sex %>%
  group_by(sex, age, indicator) %>%
  summarize(
    cases = sum(cases, na.rm = TRUE),
    .groups = "drop"
  )

hiv_age_sex_summary
```

```{r}
hiv_age_sex_summary%>%
  filter(indicator=="HIV diagnoses")%>%
  ggplot(aes(x = age, y = cases)) +
  geom_col() +
  facet_wrap(~ sex) +
  labs(
    title = "HIV cases by age and sex, 2022",
    x = "Age group",
    y = "Cases"
  )
```

## Comparing years

You can also query multiple years and compare trends over time.

```{r}
syphilis_years <- get_atlas(
  disease   = "adult syphilis",
  geography = "state",
  year      = 2018:2022
)

head(syphilis_years)
```

For example, compare national totals by year.

```{r}
syphilis_trend <- syphilis_years %>%
  group_by(year, indicator) %>%
  summarize(
    cases = sum(cases, na.rm = TRUE),
    .groups = "drop"
  )

syphilis_trend
```

```{r}
syphilis_trend%>%
ggplot(mapping= aes(x = year, y = cases, color=indicator, group=indicator)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Adult syphilis cases over time",
    x = "Year",
    y = "Cases"
  )
```

## Working with censored or missing values

CDC AtlasPlus may suppress or censor some values, especially for small counts.
Depending on the disease and geography, `cases` or `rate100000` may be missing.

A common first step is to inspect missingness.

```{r}
chlamydia_county %>%
  summarize(
    n_rows = n(),
    missing_cases = sum(is.na(cases)),
    missing_rates = sum(is.na(rate100000))
  )
```


## Choosing diseases and stratification variables

The available diseases and stratification variables depend on CDC AtlasPlus.
Common disease values include:

```{r, eval=FALSE}
get_atlas(disease = "hiv", geography = "state", year = 2022)
get_atlas(disease = "chlamydia", geography = "state", year = 2022)
get_atlas(disease = "gonorrhea", geography = "state", year = 2022)
get_atlas(disease = "adult syphilis", geography = "state", year = 2022)
get_atlas(disease = "tuberculosis", geography = "state", year = 2022)
```


## Next steps

After getting comfortable with basic queries, consider:

- querying multiple diseases
- comparing state and county results
- using sex, age, and race/ethnicity stratification
- joining AtlasPlus data to Census or spatial data
- using advanced vignettes for tract-level extrapolation and geographic access analyses

See `?get_atlas` for full argument documentation.
